---
title: "De novo assembly Day 2"

engine: knitr 

execute:
    eval: false

format:
    html:
        toc: true
        embed-resources: true
        theme: cosmo
        css: config/styles.css
        link-external-newwindow: true

filters:
    - config/italics-denovo.lua
    - config/links.lua

bibliography: config/references.bib
---

## Today's aims

In the first part of this tutorial you familiarized yourself with the data, performed quality filtering and started a de novo assembly with Trinity. Today, you will:

-   Assess the quality of the assembly
-   Perform gene prediction and annotation
-   Identify differentially expressed genes

## Assess the assembly quality

De novo transcriptome assemblies are not error free. The assembler and the parameters you choose can affect the final results. That is why it is important to assess the quality of your assembly before moving to downstream analyses.

Unlike reference-guided methods, evaluating a *de novo* assembly can be challenging since we lack a reference genome. However, there are different metrics that can be used and many can be computed with helper scripts included in Trinity. Some key metrics:

-   **Fragmentation and contig length**. If your assembly contains many short contigs (less than 200 bp) it may be very fragmented. Ideally, you want fewer, longer contigs.
-   **Read support/Read mapping rate**. This metric measures how well the original reads map back to the assembled transcripts. A good assembly should have \>80% of reads mapping back to the assembly.
-   **Transcriptome completeness**. This metric measures how complete your assembly is. You can assess this by checking how many conserved, single-copy genes (orthologues) are present in the assembly. This can be calculated with tools like BUSCO. A good assembly generally falls in the 80-90% completeness range, though the exact cutoff depends on how many marker genes were used.

### Cleanup and getting basic statistics

```{bash}
# Check in the slurm log file whether the Trinity run finished successfully
# If trinity finished successfully you should see:
# `Finished. Final Trinity assemblies are written to ...`
# If it did not finish, then you can copy an assembly from here:
# /zfs/omics/projects/education/BBDA_transcriptomics/denovo/trinity/...

# Check if --full-cleanup worked, 
# you only want the following files in the trinity output folder:
# trinity.Trinity.fasta, trinity.Trinity.fasta.gene_trans_map
# If there are more files in the folder, run the following two commands
rm -r trinity/trinity/chrysalis/ trinity/trinity/insilico_read_normalization/ trinity/trinity/read_partitions/ trinity/trinity/__salmon_filt.chkpts/ trinity/trinity/salmon_outdir/ trinity/Trinity.tmp.fasta trinity/Trinity.tmp.fasta.salmon.idx/

rm trinity/trinity/*.fa trinity/trinity/*.ok trinity/trinity/*.sam trinity/trinity/*finished trinity/trinity/*list trinity/trinity/*cmds trinity/trinity/*histo trinity/trinity/*completed trinity/trinity/.inchworm* trinity/trinity/.jellyfish* trinity/trinity/.iworm* trinity/trinity/*_count

# Check the trinity documentation and check whether there is 
# a helper script that can be used to get summary statistics for the assembly
# Hint: This helper script runs very fast, feel free to run it on the login node
conda activate /zfs/omics/projects/education/BBDA_transcriptomics/denovo/envs/trinity_bbda
...
```

::: {.callout-caution collapse="true" appearance="simple"}
## Code example

```{bash}
# Check in the slurm log file whether the Trinity run finished successfully
tail logs/trinity_*.out

# Optionally clean up the trinity folder
rm -r trinity/trinity/chrysalis/ trinity/trinity/insilico_read_normalization/ trinity/trinity/read_partitions/ trinity/trinity/__salmon_filt.chkpts/ trinity/trinity/salmon_outdir/ trinity/Trinity.tmp.fasta trinity/Trinity.tmp.fasta.salmon.idx/

rm trinity/trinity/*.fa trinity/trinity/*.ok trinity/trinity/*.sam trinity/trinity/*finished trinity/trinity/*list trinity/trinity/*cmds trinity/trinity/*histo trinity/trinity/*completed trinity/trinity/.inchworm* trinity/trinity/.jellyfish* trinity/trinity/.iworm* trinity/trinity/*_count

# Generate summary statistics
conda activate /zfs/omics/projects/education/BBDA_transcriptomics/denovo/envs/trinity_bbda

TrinityStats.pl \
    trinity/trinity.Trinity.fasta \
    >  trinity/trinity_stats.txt

conda deactivate
```
:::

::: {.callout-note title="Question" appearance="simple"}
1.  Compare your assembly statistics to those reported in the Ruocco et al. (2017) paper. Focus on:
    1.  The total number of transcripts
    2.  The N50 value

How does your assembly compare to the numbers reported in the paper?

::: {.callout-tip collapse="true" appearance="simple"}
## Answer

1.  Comparison to the paper:
    1.  Your assembly likely produced  ~80,000 transcripts, which is less than the \~170,000 reported in the original study. The difference is expected because we worked with subsampled reads to reduce the computational load. This is a good example for how reducing sequencing depth might result in an under-representation of lowly expressed transcripts.
    2.  The N50 value should be around 2,000 bp, which suggests the assembly quality is in line with the original study despite having fewer input reads.

🔎 **Good to Know**

The N50 represents the length of the shortest contig such that 50% of the total assembled sequence is contained in contigs of that length or longer. For a genome assembly, the goal is to have the highest possible N50 value as this indicates a more complete assembly. For transcriptome assemblies, however, the N50 can be misleading. That is because a transcriptome contains many short transcripts that might skew this metric. Because of that, you should interpret the N50 not by itself but together with other metrics.

Your trinity statistics should look something like this:

```         
################################
## Counts of transcripts, etc.
################################
Total trinity 'genes':  37356
Total trinity transcripts:      77264
Percent GC: 43.20

########################################
Stats based on ALL transcript contigs:
########################################

        Contig N10: 4453
        Contig N20: 3449
        Contig N30: 2866
        Contig N40: 2446
        Contig N50: 2095

        Median contig length: 1066
        Average contig: 1376.12
        Total assembled bases: 106324641


#####################################################
## Stats based on ONLY LONGEST ISOFORM per 'GENE':
#####################################################

        Contig N10: 4431
        Contig N20: 3419
        Contig N30: 2799
        Contig N40: 2327
        Contig N50: 1918

        Median contig length: 466
        Average contig: 972.96
        Total assembled bases: 36345773
```
:::
:::

### Assess transcriptome completeness

::: {.callout-important collapse="false" appearance="simple"}
**Do not yet run this section.**

Since we do have a lot of things planned for today, only work on this section once you have finished everything else.
:::

[BUSCO](https://busco.ezlab.org/busco_userguide.html) (Benchmarking Universal Single-Copy Orthologs) is a tool to assess the completeness of genome or transcriptome assemblies [@Simão2015]. It does so by searching for a set of highly conserved single-copy genes that should be present in nearly all members of a given taxonomic group.

To pick a good marker gene set, first look up where your species sits in the taxonomic tree, for example with the [NCBI Taxonomy Browser](https://ncbi.nlm.nih.gov/Taxonomy/Browser/wwwtax.cgi), and then check which lineages BUSCO actually has a marker set for (run `busco --list-datasets` to see all of them). NCBI places *Cymodocea nodosa* under Liliopsida (the monocots), inside the order Alismatales and family Cymodoceaceae. BUSCO does not offer a marker set for Alismatales or Cymodoceaceae, so `liliopsida_odb12.2` is the most specific, and therefore most informative, marker set currently available for this species.

```{bash}
# Create an output directory
...

# Run BUSCO  
# Estimated run time: ~ 20 min
# Hints:    - Ensure that you use BUSCO in a mode that is suited for transcriptomes 
#           - Ensure that it is run with the marker gene set that best matches
#             our species (i.e. use liliopsida_odb12.2, see the paragraph above
#             for why this is the best available choice)
#           - BUSCO can be accessed with the `transcriptomics_bbda` conda environment
#           - Use 10 CPUs and 50G of memory
...
```

::: {.callout-caution collapse="true" appearance="simple"}
## Code example

```{bash}
# Create an output directory
mkdir trinity/busco

# Run BUSCO 
conda activate /zfs/omics/projects/education/BBDA_transcriptomics/denovo/envs/transcriptomics_bbda 

srun --cpus-per-task 10 --mem=50G busco \
    -i trinity/trinity.Trinity.fasta \
    -m transcriptome -f \
    -c 10 --out trinity/busco \
    -l liliopsida_odb12.2 

conda deactivate
```
:::

::: {.callout-note title="Question" appearance="simple"}
After BUSCO completes, answer the following:

1.  How complete is the assembly?
    -   Look up how BUSCO defines completeness.
    -   Are there limitations to using BUSCO to assess the completeness of transcriptome assemblies?
2.  How many duplicates are there?
    -   Does this number seem unusually high?
    -   Should we be concerned about this number?

::: {.callout-tip collapse="true" appearance="simple"}
## Answer

**1. Assembly completeness**

Your BUSCO output will look something like this: `C:81.5%[S:19.0%,D:62.5%],F:3.6%,M:15.0%,n:3322`.

This tells that BUSCO searched for 3322 marker genes and of these:

-   81.5% were present and of these 19.0% were single-copy and 62.5% appeared as duplicates
-   3.6% of markers were fragmented
-   15.0% of markers were missing

Keep in mind that this is a small fraction of your assembly: BUSCO only evaluates these 3322 marker genes, compared to the ~37,000 trinity 'genes' (and ~77,000 transcripts) reported by TrinityStats. A high completeness score tells you that this specific, conserved set of genes is present, not that the rest of the several-thousand-gene assembly is complete or accurate.

A transcriptome completeness score in the 80-90% range is generally considered good, though the exact cutoff depends on how many marker genes were used. BUSCO assesses completeness by looking for a lineage-specific number of marker genes, and this number varies a lot between lineages. In our case, these are 3322 genes that are specific for Liliopsida (monocots), the most specific marker set currently available for our species. A larger, more specific marker set is more likely to include genes that are lowly expressed or not expressed at all under our experimental conditions, which is also why the completeness score here might be a bit lower, and the missing fraction higher, than you would see with a smaller, more general marker set like eukaryota, which has 255 marker genes.

Marker genes are selected as those that are present in at least 90% of the species in a given lineage, and present in a single copy in 90% of those species. These genes are mostly housekeeping genes, such as ribosomal proteins, which are usually highly expressed and often easier to assemble. Therefore, high BUSCO scores do not guarantee that all relevant or condition-specific genes were captured. Genes that are poorly expressed or not expressed at all under your experimental conditions may still be missing from the assembly.

**2. Duplication rate**

Around 62.5% of the detected marker genes were duplicates, which might seem high. However, since we work with a transcriptome a single gene can often produce multiple isoforms, which appear as duplicated entries to BUSCO. This is expected and we therefore do not need to worry about a high duplication rate when working with transcriptomes.
:::
:::

## Annotation

After assembly and quality control, we will translate the transcript sequences into coding sequences (CDS), identify proteins and then annotate their biological function. In the next section you will do the following:

-   Perform gene calling using TransDecoder
-   Do a functional annotation using HMMER with the KEGG database

### Gene calling

We will use [TransDecoder](https://github.com/TransDecoder/TransDecoder/wiki) to identify likely coding regions (open reading frames, ORFs) and predict proteins from our transcriptome assembly [@haas2025].

```{bash}
# Prepare an output folder
...

# Run TransDecoder
# Estimated run time: ~ 20 min
# Hints:  - Transdecoder can be accessed via the `transcriptomics_bbda` environment
#         - TransDecoder now runs as a single command that internally finds long
#           open reading frames (ORFs) first, then predicts which ones are likely
#           real coding regions
#         - Ensure that you account for strand information
#         - Make use of the gene to transcript map
#         - Ensure that you only produce a single output cds sequence/transcript to simplify our analysis
#         - Use srun with 1 CPU and 20G of memory
...

# Use cut to simplify the header of the protein sequences and to 
# remove everything after the space.
# That is a useful habit in order to keep protein (and names in general) short and simple
# It is also useful if you want to, for example, parse annotations further down the line
...
```

::: {.callout-caution collapse="true" appearance="simple"}
## Code example

```{bash}
# Prepare an output folder
mkdir annotations

# Run transdecoder 
conda activate /zfs/omics/projects/education/BBDA_transcriptomics/denovo/envs/transcriptomics_bbda

srun --cpus-per-task 1 --mem=20G TransDecoder -t trinity/trinity.Trinity.fasta \
    --gene_trans_map trinity/trinity.Trinity.fasta.gene_trans_map \
    -S \
    --single_best_only \
    --output_dir annotations

# Simplify the header
cut -f1 -d " " annotations/trinity.Trinity.fasta.transdecoder.pep > annotations/Trinity.fasta.transdecoder_short.pep

conda deactivate
```
:::

### Functional annotation

Next, we will functionally annotate the predicted proteins by comparing them to the [KEGG Orthology (KO) database](https://www.genome.jp/kegg/pathway.html) using HMMER's [hmmsearch algorithm](http://eddylab.org/software/hmmer/Userguide.pdf)[@Finn2015; @Kanehisa2000]. This will allow us to link each transcript to a potential function.

Note that this will run for 1-2 hours. So once the job started successfully, you can proceed with the following steps. You only will need the results of this analysis once you finished the DEG analysis.

```{bash}
# Write a sbatch script to run a hmmsearch to compare your proteins against KEGG profiles
# Estimated run time: 1-2 hours
# Hints:  - Use the pep file with the simplified header, otherwise you will run into issues with the following steps!
#         - The KEGG database can be found at 
#           `/zfs/omics/projects/education/BBDA_transcriptomics/denovo/databases/KO_db.hmm`
#         - Use a setting to generate a table of per-sequence hits
#         - add this setting: --notextw 
#         - use at max 10 CPUs, 20G of memory and 4 hours of run time
mkdir annotations/KEGG/
...

# Filter the HMMER results to keep only the best hit per transcript
# The code below does the following (and might take a minute or so to finish):
# Step 1: Convert consecutive spaces to tabs 
# Step 2: Remove lines starting with "#"
# Step 3: Extract columns of interest: transcript ID, target name, bitscore, and e-value
# Step 4: Keep only hits with e-value <= 1E-3
# Step 5: Sort first by transcript ID, then by highest bitscore and lowest e-value
# Step 6: From the sorted list, keep only the top hit per transcript
sed 's/ \+ /\t/g' annotations/KEGG/sequence_results.txt | \
    sed '/^#/d'| sed 's/ /\t/g'| awk -F'\t' -v OFS='\t' '{print $1, $3, $6, $5}' | \
    awk -F'\t' -v OFS='\t' '($4 + 0) <= 1E-3' | \
    sort -t$'\t' -k3,3gr -k4,4g | sort -t$'\t' --stable -u -k1,1 \
    >  annotations/KEGG/sequence_results_best_hit.txt

# Merge the filtered HMMER results with the KEGG Orthology (KO) mapping file
# Step 1: Inside `<()` sort both files to prepare them for 'join'
#         - sequence_results_best_hit.txt is sorted by column 2 (KO ID)
#         - ko_list is sorted by column 1 (KO ID)
# Step 2: Use 'join' to combine the two files on the KO ID columns
#         - -1 2 : match column 2 of the first file
#         - -2 1 : match column 1 of the second file
#         - -a1 : keep all rows from the first file, even if no match is found
#         - -e '-' : fill in missing values with '-'
#         - -t $'\t' : use tab as the delimiter
#         - -o ... : specify output order 
#                    (from file 1: cols 1,2,4,3; from file 2: cols 2 and 12)
# Step 3: Sort the merged output and save as 'temp'
LC_ALL=C join -a1 -1 2 -2 1 -e'-' -t $'\t' -o1.1,1.2,1.4,1.3,2.2,2.12 <(LC_ALL=C sort -k2 annotations/KEGG/sequence_results_best_hit.txt) <(LC_ALL=C sort -k1  /zfs/omics/projects/education/BBDA_transcriptomics/denovo/databases/ko_list) | LC_ALL=C  sort >  temp

# Add a header line to the final output file
# Step 1: Use 'echo' to print the header with column names
# Step 2: Use 'cat - temp' to concatenate the header (-) with the contents of 'temp'
#         - here, '-' tells 'cat' to read from standard input (the echoed header)
#         - 'temp' is the file with our merged results
# Step 3: Save everything into the final output file 'annotations/KEGG/annotations.txt'
echo -e "accession\tKO_hmm\tKO_e_value\tKO_bit_score\tKO_bit_score_cutoff\tKO_Definition" | cat - temp > annotations/KEGG/sequence_results_best_hit_desc.txt

# Cleanup intermediate files
rm temp 
```

::: {.callout-caution collapse="true" appearance="simple"}
## Code example

```{bash}
# Write a sbatch script to run a hmmsearch to compare proteins against KEGG profiles
mkdir annotations/KEGG/
sbatch scripts/hmmsearch_kegg.sh 

'''
#!/bin/bash
#SBATCH --job-name=hmmer
#SBATCH --output=logs/hmmer_%j.out
#SBATCH --error=logs/hmmer_%j.err
#SBATCH --cpus-per-task=10
#SBATCH --mem=20G
#SBATCH --time=02:00:00

hmmsearch \
    --tblout annotations/KEGG/sequence_results.txt \
    --notextw \
    --cpu 10 \
    /zfs/omics/projects/education/BBDA_transcriptomics/denovo/databases/KO_db.hmm \
    annotations/Trinity.fasta.transdecoder_short.pep
'''

# Filter the HMMER results to keep only the best hit per transcript
sed 's/ \+ /\t/g' annotations/KEGG/sequence_results.txt | \
    sed '/^#/d'| sed 's/ /\t/g'| awk -F'\t' -v OFS='\t' '{print $1, $3, $6, $5}' | \
    awk -F'\t' -v OFS='\t' '($4 + 0) <= 1E-3' | \
    sort -t$'\t' -k3,3gr -k4,4g | sort -t$'\t' --stable -u -k1,1 \
    >  annotations/KEGG/sequence_results_best_hit.txt

# Merge the filtered HMMER results with the KEGG Orthology (KO) mapping file
LC_ALL=C join -a1 -1 2 -2 1 -e'-' -t $'\t' -o1.1,1.2,1.4,1.3,2.2,2.12 <(LC_ALL=C sort -k2 annotations/KEGG/sequence_results_best_hit.txt) <(LC_ALL=C sort -k1  /zfs/omics/projects/education/BBDA_transcriptomics/denovo/databases/ko_list) | LC_ALL=C  sort >  temp

# Add a header line to the final output file
echo -e "accession\tKO_hmm\tKO_e_value\tKO_bit_score\tKO_bit_score_cutoff\tKO_Definition" | cat - temp > annotations/KEGG/sequence_results_best_hit_desc.txt

# Cleanup intermediate files
rm temp 
```
:::

::: {.callout-note title="Question" appearance="simple"}
1.  Inspect the TransDecoder output
    1.  How many predicted proteins are there?
    2.  How does this compare to the number of input transcripts? Why might these numbers differ?
2.  Check the output of the KEGG annotation:
    1.  How many proteins were assigned to a KO?
    2.  How many unique KOs were found?

::: {.callout-tip collapse="true" appearance="simple"}
## Answer

1.  Transdecoder output:
    1.  We started with \~77,000 transcripts and ended up with \~44,000 proteins
    2.  We have fewer proteins than transcripts because
        1.  Not all transcripts contain coding sequences (CDS)
        2.  Short transcripts, potential artifacts (e.g., chimeras), or untranslated regions will not result in a protein sequence
2.  KEGG annotation output:
    1.  We work with \~44,000 proteins and of these \~35,000 should be assigned to a KO (\~80%)
    2.  We work with \~5,000 unique KOs

```{bash}
# Question 1.1:
# We work with 44,638 proteins and 77,264 transcripts
grep -c ">" annotations/Trinity.fasta.transdecoder_short.pep
grep -c ">" trinity/trinity.Trinity.fasta

# Question 2.1:
# Of all proteins 35,571 are assigned to a KO 
wc -l annotations/KEGG/sequence_results_best_hit.txt 

# Question 2.2
# We work with 5094 unique KOs, i.e. ortholog groups 
cut -f2 annotations/KEGG/sequence_results_best_hit.txt | sort | uniq | wc -l
```
:::
:::

## Identifying differentially expressed genes

There are several methods for estimating transcript abundance in a genome-free manner, and these include alignment-based methods (aligning reads to the transcript assembly) and alignment-free methods (typically examining k-mer abundances in the reads and in the resulting assemblies).

In this section, we will make use of a Trinity helper script to quantify transcript and gene-level abundances of each of our 6 samples using the alignment-free method salmon [@Patro2017].

### Quantification with salmon

```{bash}
# Write an sbatch script to run salmon for transcript quantification
# Estimated run time: ~ 5 min
# Hints:    - Check the trinity documentation and find a perl helper script that can be used to quantify with salmon
#           - Submit this via sbatch with 4 CPUs, 20G of memory and a 2h hour run time 
#           - Ensure that you activate the transcriptomics_bbda conda environment
#           - Do not forget about strand orientation 
#           - Use `--gene_trans_map  xxx/Trinity.fasta.gene_trans_map` to quantify on transcript as well as gene level
#           - Add `--prep_reference` to build a target index
...

# Cleanup:
# unfortunately, the --output_dir does not work very well and 
# Salmon will generate one output folder per sample in the working dir, 
# therefore, use the mv command to generate an organized folder structure
...

# The salmon log file can be used to extract the mapping rate, 
# another useful statistic to assess the assembly quality.
# We provide a small helper script for this.
# Use it and check what the mapping rate is
cp /zfs/omics/projects/education/BBDA_transcriptomics/denovo/scripts/parse_salmon.py scripts 
python scripts/parse_salmon.py -i logs/salmon_*.err

# With this you have all files for your DEG analysis with DESeq in R 
# You can do the analysis on Crunchomics or on your own computer if you prefer to use RStudio 
# If you want to do the later you need to transfer the salmon folders, 
# assembly, gene to transcript mapping and parsed annotations
# for the downstream analyses to your own computer
scp ... 

# Hint: You can use rsync to download and maintain a folder structure 
# but have fine-grained control over what you want to access 
# Step 1:  -a : archive mode (preserves permissions, timestamps, symlinks)
#         -v : verbose (prints progress)
# Step 2: Keep directory structure but only include specific files
#         --prune-empty-dirs : remove any empty directories from the copy
#         --include='*/'     : include all directories so structure is preserved
#         --include='quant.sf' : include transcript-level quantification files
#         --include='quant.sf.genes' : include gene-level quantification files
#         --exclude='*'      : exclude everything else
rsync -av --prune-empty-dirs \
    --include='*/' \
    --include='quant.sf' \
    --include='quant.sf.genes' \
    --exclude='*' \
    crunchomics:/path/salmon_abund/ results/trinity/salmon_abund/
```

::: {.callout-caution collapse="true" appearance="simple"}
## Code example

```{bash}
# Write an sbatch script to run salmon for transcript quantification

sbatch scripts/run_salmon.sh

'''
#!/bin/bash
#SBATCH --job-name=salmon
#SBATCH --output=logs/salmon_%j.out
#SBATCH --error=logs/salmon_%j.err
#SBATCH --cpus-per-task=4
#SBATCH --mem=20G
#SBATCH --time=02:00:00

# Activate conda environment
source ~/.bashrc
conda activate /zfs/omics/projects/education/BBDA_transcriptomics/denovo/envs/trinity_bbda

# Run Salmon
align_and_estimate_abundance.pl \
        --seqType fq \
        --samples_file files/trinity_mapping.txt \
        --transcripts trinity/trinity.Trinity.fasta \
        --SS_lib_type RF \
        --est_method salmon \
        --gene_trans_map trinity/trinity.Trinity.fasta.gene_trans_map \
        --prep_reference \
        --thread_count 4 \
        --output_dir trinity/salmon_abund/ 
'''

# Cleanup:
mkdir trinity/salmon_abund/ 
mv treated_rep* trinity/salmon_abund/
mv untreated_rep* trinity/salmon_abund/

# The salmon log file can be used to extract the mapping rate
cp /zfs/omics/projects/education/BBDA_transcriptomics/denovo/scripts/parse_salmon.py scripts 
python scripts/parse_salmon.py -i logs/salmon_*.err

# Download the data
# The rsync command will only download the quant files but keep the folder structure intact
scp user@omics-h0.science.uva.nl:/path/trinity.Trinity* results/trinity 
scp user@omics-h0.science.uva.nl:/path/annotations/KEGG/sequence_results_best_hit_desc.txt results/annotations/

rsync -av --prune-empty-dirs \
    --include='*/' \
    --include='quant.sf' \
    --include='quant.sf.genes' \
    --exclude='*' \
    user@omics-h0.science.uva.nl:/path/salmon_abund/ results/trinity/salmon_abund/
```
:::

### DEG analysis with DESeq2

After transcriptome assembly and annotation, differential gene expression (DGE) analysis helps us identify genes with significantly different expression levels between experimental conditions. Since you have worked with DESeq2 before, you can use your notes from that tutorial to run the next steps. Briefly, you want to do the following:

1.  Load the required libraries (tidyverse, tximport and DESeq2). Note, none of these are installed but you can install them yourself on your personal PC or the HPC with `install.packages("name")`
2.  Read in the salmon data with [tximport](https://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#transcript-abundance-files-and-tximport-tximeta)
    1.  You can use the salmon folder names to make a minimal samples mapping file:

    ```{r}
    # Get a list of the salmon folder names
    dir <- "results/trinity/salmon_abund"
    folders <- list.dirs(dir, full.names = FALSE, recursive = FALSE)

    # Using the folder names, extract the names of the treatments and build a samples dataframe
    condition <- ifelse(grepl("^treated", folders), "treated", "untreated")

    samples <- data.frame(
    run = folders,
    condition = factor(condition)
    )

    samples
    ```

    2.  tximport requires the tx2gene argument to summarize the data on gene level. For this we can use the gene_trans_map from Trinity. However, read the tximport documentation carefully to ensure that the format is correct. Hint: Watch the order of the columns
3.  Optional: Read in the annotations and summarize the isoform information on gene level by only selecting the "best" annotation per isform using the highest bit-score. Hint: Beware of ties.
4.  Generate a DESeqDataSet object
    1.  Ensure that the untreated samples are used as reference level
    2.  Optional: Filter low abundance hits. For example, you can require that genes have Y or more samples with counts of X or more
5.  Optional: Visualize the data with a PCA. This is a good sanity check to ensure that the replicates cluster as expected.
6.  Identify differentially expressed genes. Hint: In the `results()` function, use reasonable cutoffs for alpha and lfcThreshold to focus on "high-confidence" DEGs
7.  Add the information from the annotations to the results to be able to quickly screen the list for genes of interest.

::: {.callout-caution collapse="true" appearance="simple"}
## Code example

### General setup

```{r}
#| eval: true
#| warning: false
#| message: false
# Load necessary libraries 
library("tidyverse")
library("tximport")
library("DESeq2")
```

### Read in salmon data

```{r}
#| eval: true
#| warning: false
#| message: false
# Locate the directory containing the salmon files
dir <- "results/trinity/salmon_abund"
folders <- list.dirs(dir, full.names = FALSE, recursive = FALSE)
folders 

# Using the folder names:
# Generate a data frame that contains the new sample names, i.e. treatment_rep1,
# and the treatment (treated versus untreated)
condition <- ifelse(grepl("^treated", folders), "treated", "untreated")
condition 

samples <- data.frame(
    run = folders,
    condition = factor(condition)
)

samples

# Create a named vector that points to the sample files
files <- file.path(dir, samples$run, "quant.sf")
names(files) <- samples$run
files 

# For gene-level summarization link the transcript to gene IDs 
# For this tximport needs a dataframe with two columns: transcript ID and gene ID 
# Trinity provides this file but in the wrong order, so we have to clean it up first
## We have ~77,264 transcripts, as expected
tx2gene <- read_tsv("results/trinity/trinity.Trinity.fasta.gene_trans_map",  col_names = FALSE)
tx2gene <- tx2gene[,c(2,1)]   # reverse the order of the columns
colnames(tx2gene) <- c("transcript", "gene") # add a sensible header
str(tx2gene)

# Read in the files using tximport 
## We imported 37356 "genes" from 6 samples
txi <- tximport(files, type = "salmon", tx2gene = as.data.frame(tx2gene))
str(txi)
```

### Read in and summarize annotations

```{r}
#| eval: true
#| warning: false
#| message: false
# Read in the annotation data
annotations_df <- read_tsv("results/annotations/sequence_results_best_hit_desc.txt")

# Since several genes can consist of more than one transcript, we will now condense the data by gene (by choosing the top hit)
annotations_gene_level <- annotations_df |>
    mutate(
        # Extract gene id by ignoring the _i.* part of the transcript name
        gene = sub("_i.*", "", accession)
    ) |>
    group_by(gene) |>
    arrange(desc(KO_bit_score), KO_e_value) |>
    slice_head(n = 1) |>
    ungroup()

# Compare the df sizes 
## Initial df: 35571 (as expected)
## Df on gene level: 11049 (we see to have quite some isoforms)
dim(annotations_df)
dim(annotations_gene_level)
```

### Generate a DESeqDataSet object

```{r}
#| eval: true
#| warning: false
#| message: false
# Generate a DESeqDataSet object
## The object has 37356 rows and 6 cols as expected for the data input
dds = DESeqDataSetFromTximport(txi, colData = samples, design = ~condition)
dds 

# Remove uninformative columns (optional)
# This step only keeps genes that have Y or more samples with counts of X or more (keep <- rowSums( counts(dds) >= X ) >= Y)
# Choose: Y = 3 because thats the nr of replicates, and 3 a commonly used cutoff
# I used 3 here, since we work with down-sampled data
## The object now has 20544 rows and colums suggesting a large proportion of transcripts are lowly expressed
keep <- rowSums( counts(dds) >= 3 ) >= 3
dds <- dds[keep, ]
dds 

# Ensure that untreated samples are seen as the reference level
dds$condition <- relevel(dds$condition, ref = "untreated")
```

### Perform quality assessment

```{r}
#| eval: true
#| warning: false
#| message: false
# Perform rlog transformation for explorative data visualization 
rld_blindTRUE <- rlog(dds, blind=TRUE)

# Generate a PCA plot 
pcaData <- plotPCA(rld_blindTRUE, intgroup= c("condition"), returnData = TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))

PCA_plot <- 
    ggplot(pcaData, aes(x = PC1, y = PC2)) +
        geom_point(size = 3, aes( color = condition), alpha = 0.7) +
        xlab(paste0("PC1: ", percentVar[1], "% variance")) +
        ylab(paste0("PC2: ", percentVar[2], "% variance")) +
        labs(caption = "Plotted data using blind=TRUE") +
        theme_classic()

PCA_plot
```

Note: Replicates don't cluster cleanly together but that can be expected for this down-sampled set.  We still see that untreated is closer to each other than to the treated samples. We also see that the treated samples are more variable which might also be expected considering we have a strong treatment that might have various effects.


### Identify DEGs

```{r}
#| eval: true
#| warning: false
#| message: false
# Run DESEQ
dds <- DESeq(dds)

# Extract DEGs
res_table <- results(dds, contrast=c("condition", "treated", "untreated"), alpha = 0.05, lfcThreshold = 0.58)

# Apply log2-shrinkage 
res_table_unshrunken <- res_table
res_table <- lfcShrink(dds, contrast=c("condition", "treated", "untreated"), res = res_table, type="ashr")

# Summarize the results 
summary(res_table, alpha = 0.05)

# Convert table to tibble
res_table_tb <- res_table |> 
    data.frame() |> 
    rownames_to_column(var = "gene") |> 
    as_tibble()

# Subset DEGs only
res_table_sig <- res_table_tb |>
    filter(padj < 0.05)

# Add annotation info 
res_table_sig <- merge(res_table_sig, annotations_gene_level, by = "gene", all.x = TRUE)

# Write to file 
write.csv(res_table_sig, "results/trinity/degs.csv", row.names = FALSE)
```
:::

::: {.callout-note title="Question" appearance="simple"}
1.  How many genes were identified as DEGs. How many genes were up- and down-regulated? Is this comparable to what has been described in the paper?
2.  Look at the gene descriptions (based on the KO annotations), what kind of genes are typically differentially expressed? Again, compare this to what has been described in the paper and whether your results are consistent.
    1.  Hint: You can upload the KOs of DEGs [here](https://www.genome.jp/kegg/mapper/search.html) to get a quick glimpse at the associated metabolic pathways.
    2.  Hint: If you uploaded the KOs to the website above you might see a number of pathways related to medical pathways. You can ignore those because they are not relevant for our study system and they often tend to come up if housekeeping genes, such as ribosomal proteins, show a differential expression.

::: {.callout-tip collapse="true" appearance="simple"}
## Answer

1.  You should have found roughly 70-90 DEGs, the large majority up-regulated and very few (if any) down-regulated. Don't worry if your exact numbers differ a bit from these, small differences in the assembly or filtering choices can shift the count. We find far fewer DEGs than described in the paper (153 upregulated and 17 downregulated), likely because of the massive downsampling of reads.
2.  DEGs are associated with photosynthesis, including light reaction functions and CO2 fixation, but also to other metabolic pathways, specifically for enzymes involved in glycolysis, the tricarboxylic acid cycle and energy metabolism of the mitochondrial electron transport. This is very consistent with the paper.
:::
:::

🔎 **Good to Know**:

**Versions matter**: When running this tutorial I noticed that knowing the version of your software is crucial if you want to repeat the code as this can impact your results. This is what happened to me, when running the code in two different R environments:

| Metric              | R 4.2.2 + DESeq2 1.38.0 |  R 4.4.2 + DESeq2 1.46.0|
|---------------------|-------------------------|-------------------------|
| Significant DEGs    | 33 genes                | 58 genes                |
| Low-count filtering | 3782 genes filtered     | 2219 genes filtered     |
| Filter threshold    | mean count \< 11        | mean count \< 7         |

Don't worry if your own DEG count doesn't match any of the numbers above exactly, including the answer key further up. The point isn't the specific number, but that small differences in software versions or filtering choices can shift the result by tens of genes.

The `filterThreshold` is an automatically calculated cutoff used in DESeq2's independent filtering. It represents the minimum mean normalized count that a gene should have to be included in multiple testing correction. You can check the used threshold with `metadata(res)$filterThreshold`.

So when writing code:

-   always record the package versions in your method section
-   use `sessionInfo()` in your R scripts to automatically print all attached packages and package version


## References

::: {#refs}
:::
